---
title: "Manuscript Markdown — Revised (Reviewer Comments #1–3)"
author: "Madison Sandquist"
date: "2026-04-16"
output: github_document
---

<!--
=====================================================================
REVISION NOTES — this file implements peer-review Major Comments #1-3.
All original code is preserved; new/edited analysis is clearly marked
with headers of the form "### [REV #x] ...".

  #1 Multiple comparisons  -> Benjamini-Hochberg FDR correction added to
     the monthly correlation screen; contiguous significant blocks flagged.
  #2 Temporal autocorrelation -> Pyper & Peterman (1998) effective sample
     size (modified Chelton) used to recompute correlation p-values.
  #3 Model-selection ("forking paths") inflation -> full distribution of
     r-squared across ALL candidate windows reported per driver, number of
     candidates searched is stated, and LOO predictive R-squared is added
     to contrast in-sample fit with out-of-sample skill.

Reference for #2:
  Pyper, B.J. & Peterman, R.M. (1998). Comparison of methods to account
  for autocorrelation in correlation analyses of fish data.
  Can. J. Fish. Aquat. Sci. 55: 2127-2140.
=====================================================================
-->

<!--
NOTE: base R only is used for the new statistics (p.adjust, acf, pt, lm),
so no new package installs are required beyond what the original file loads.
-->


```{r setup, include=FALSE}
knitr::opts_knit$set(
  root.dir = "C:/Users/msand/OneDrive/Documents/GitHub/Oceanographic-drivers-of-gopher-rockfish-recruitment"
)
```

### Libraries Needed

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(car)
library(broom)
library(lme4)
library(lmerTest)
library(DHARMa)
library(emmeans) #pairwise comparisons
library(influence.ME) #outliers
library("wesanderson")
#remotes::install_github("r4ss/r4ss")
library(r4ss)
library(zoo)
library(patchwork)
library(stringr)
library(pls)
library(plsVarSel)
library(gridExtra)
library(grid)
library(slider)
library(glue)
```

### Increasing font size and color for graphing
```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Set theme with larger axis titles and numbers
larger_axis_theme <-
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black"),
    legend.position = "none",
    axis.title = element_text(size = 16),   # Axis titles
    axis.text = element_text(size = 14),
    title = element_text(size = 20)# Axis numbers
  )

colors <- wes_palette("AsteroidCity1", type = "discrete")

```

## Retrieving Stock Assessment Files
```{r, warning=FALSE}
#recommended work flow, define model directory
# stock assessment file from PFMC website from the pfmc website 
model_path <- "D:/GPHR_BYEL_POST_STAR_BASE_MODEL" # this is the stock assessment file saved on an external hard drive
goph2019 <- r4ss::SS_output(model_path, printstats = FALSE)
with.covar = T	 
parameters <- goph2019$parameters
goph_recruit <- goph2019$recruit

#not filtered for graphing 
goph_recruit_nf <-goph_recruit %>% 
  filter(Yr >= 1987, Yr <= 2018) %>% 
  rename(Year = Yr)

#filter for years to match the ROMS outputs
goph_recruit <-goph_recruit %>% 
  filter(Yr >= 1995, Yr <= 2018) %>% 
  rename(Year = Yr) 

# the pisco data from the SA file
PISCO_age0_index <- read_csv("Data/PISCO_age0_index.csv")
```

#### Figure 1 - Recruitment Deviation Time series
```{r, warning=FALSE}
# ========== Time series of recruitment devs from SA ==========
stock_assessment <- ggplot(goph_recruit_nf, aes(x = Year, y = dev)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 1)+
  geom_vline(xintercept = 1995, linetype = "solid", color = "indianred")+
  theme_classic() +
  labs(title = "", y = "Recruitment Deviation", x = "Year") + 
  larger_axis_theme
# ========== Time series of PISCO index of abundance ==========
pisco_index <- ggplot(PISCO_age0_index, aes(x = Year, y = obs)) +
  geom_line(color = "navyblue", size = 1.2) +
  theme_classic() +
  labs(title = "", y = "Index of Abundance", x = "Year") + 
  scale_x_continuous(limits = c(1987, max(PISCO_age0_index$Year))) +
  larger_axis_theme


# ========== Combined TimeSeries Plot ==========
# Then combine them
fig1 <- stock_assessment / pisco_index + plot_layout(ncol = 1)
fig1
ggsave("Figures/fig1.png", fig1, width = 7, height = 5, dpi = 300)


# ============= Corrleation plot ===========================

recruit_merged <- left_join(PISCO_age0_index, goph_recruit, by = "Year")
summary(recruit_merged)

# Fit a linear model 
lm_recruit <- lm(dev ~ obs, data = recruit_merged)
summary(lm_recruit)


recruit_plot <- ggplot(recruit_merged, aes(x = obs, y = dev)) +
  geom_point(size = 3, shape = 21, fill = "grey30") +
  geom_smooth(method = "lm", se = TRUE, size = 1.2, color = "black") +
  annotate("text", x = Inf, y = Inf, label = paste0("p < 0.001 r² = 0.87"),
           hjust = 1, vjust = 1, size = 5, fontface = "italic") +
  labs(
    x = "Index of Abundance (PISCO)",
    y = "Recruitment Deviations (Stock Assessment)",
    title = ""
  ) +
  larger_axis_theme +
  theme(axis.title.y = element_blank(), 
        theme(axis.text = element_text(size = 16)))+
  theme_classic()
recruit_plot
cor_plot <- ggsave("Figures/corr_rec.png", recruit_plot, width = 4, height = 4, dpi = 300)
```

## Cleaning and Combining Oceanographic Parameters

#### Ocean model inputs from ROMS

This data has been clean in python previous (see Python file for cleaning code)

```{r}
# load in data frame that is just DO, pH, temperature, CHLa,
df_ocean <- read_csv("Data/monthly_averages_allyear_1995_2020_ALL_41to34p4.csv")

df_ocean <- df_ocean %>% 
  select(-Chla_avg, -Chla_sd)

# load in data frame that is wind
df_wind <- read.csv("Data/monthly_upwelling_wind_offshore_averages_ALL_41to34p4.csv")

# change coding of month to merge
df_wind1 <- df_wind %>%
  mutate(Month = case_when(
    Month == "January"   ~ "Jan",
    Month == "February"  ~ "Feb",
    Month == "March"     ~ "Mar",
    Month == "April"     ~ "Apr",
    Month == "May"       ~ "May",
    Month == "June"      ~ "Jun",
    Month == "July"      ~ "Jul",
    Month == "August"    ~ "Aug",
    Month == "September" ~ "Sep",
    Month == "October"   ~ "Oct",
    Month == "November"  ~ "Nov",
    Month == "December"  ~ "Dec",
    TRUE ~ Month  # keep original if no match
  ))

# Combine into one data frame 
ROMS_outputs <- left_join(df_ocean, df_wind1, by = c("Year", "Month"))
```

#### Coastal Upwelling Transport Index

```{r}
cuti <- read.csv("Data/CUTI_daily.csv") %>%
  select(year, month, day,
         X35N, X36N, X37N, X38N, X39N, X40N, X41N) %>%
  filter(year < 2020)%>% 
  filter(year > 1994)

# 1) Monthly mean for each latitude (averaging over days)
monthly_lat <- cuti %>%
  group_by(year, month) %>%
  summarise(across(starts_with("X"), ~ mean(.x, na.rm = TRUE)), .groups = "drop") %>%
  pivot_longer(cols = starts_with("X"), names_to = "lat_col", values_to = "CUTI") %>%
  mutate(lat = as.integer(str_extract(lat_col, "\\d+"))) %>%
  filter(between(lat, 35, 41))

# 2) One monthly mean across ALL latitudes 34–41 (equal weight per latitude)
cuti_tidy <- monthly_lat %>%
  group_by(year, month) %>%
  summarise(CUTI = mean(CUTI, na.rm = TRUE), .groups = "drop") %>%
  rename(Year = year, Month = month) %>%
  mutate(
    Month = month.abb[as.integer(Month)]
  ) %>%
  arrange(Year, match(Month, month.abb))
```

#### Sea Level

```{r}
sl <- read.csv("Data/SLH_FULL.csv") %>% 
  filter(SL > 0)
  
monthly_avg_sl <- sl %>%
  group_by(Year, Month) %>%
  summarize(mean_sl = mean(SL, na.rm = TRUE), .groups = "drop") %>% 
  filter(Year >= 1995 & Year <= 2020)

```

#### PDO

```{r}
pdo <- read.csv("Data/pdo.timeseries.sstens.csv")

# Convert the Date column to Date type
pdo$Date <- as.Date(pdo$Date)

# Filter the dataframe between 1995 and 2018
monthly_avg_pdo <- pdo %>%
  filter(Date >= as.Date("1995-01-01") & Date <= as.Date("2020-12-01")) %>% 
  rename(mean_pdo = PDO.from.Ensemble.SST.https...psl.noaa.gov..pdo..Using.EOF.from.1920.to.2014.for.N.Pacific..see.webpage.) %>% 
    mutate(
    Year = year(Date),
    Month = month(Date)
  )

```

#### NPGO

```{r}
npgo <- read.csv("Data/npgo.csv")

monthly_avg_npgo <- npgo %>%
  mutate(Date = mdy(Date),
         Year = year(Date),
         Month = month(Date)) %>%
  filter(Year >= 1995 & Year <= 2020) %>%
  group_by(Year, Month) %>%
  summarise(npgo_avg = mean(NPGO, na.rm = TRUE), .groups = "drop")

```

#### ONI

```{r}
ONI <- read.csv("Data/ONI_data.csv")
monthly_avg_ONI <- ONI %>%
  pivot_longer(
    cols = all_of(month.abb),  # Ensures only "Jan" to "Dec"
    names_to = "Month_name",
    values_to = "oni_avg"
  ) %>%
  mutate(
    Month = match(Month_name, month.abb)
  ) %>%
  select(Year, Month, oni_avg) %>%
  arrange(Year, Month) %>% 
  filter(Year > 1994) %>% 
  filter(Year < 2020)

```

#### Combine indices into single data frame

```{r}
# non spatially separated 
combined_monthly_indices <- monthly_avg_sl %>%
  full_join(monthly_avg_pdo, by = c("Year", "Month")) %>%
  full_join(monthly_avg_npgo, by = c("Year", "Month")) %>%
  full_join(monthly_avg_ONI, by = c("Year", "Month")) %>% 
  arrange(Year, Month) %>%
  filter(Year < 2019) %>%
  mutate(Month = month.abb[Month])
head(combined_monthly_indices)

# Combined with ROMS with cuti 
ROMS_CUTI_environmental_combined_all_yr <- ROMS_outputs %>%
  full_join(cuti_tidy, by = c("Year", "Month"))

head(ROMS_CUTI_environmental_combined_all_yr)

df_all_parameters <- ROMS_CUTI_environmental_combined_all_yr %>% 
  full_join(combined_monthly_indices, by = c("Year", "Month"))%>% 
  select(-Date)

head(df_all_parameters)
```

### Oceanographic Time Series (Figures 2 and 3)

#### Figure 2. Nearshore Habitat From 1995 to 2018
```{r}
ROMS_CUTI_environmental_combined_all_yr_plot <- ROMS_CUTI_environmental_combined_all_yr %>%
  filter(Year <= 2018) %>% 
  mutate(
    Month_num = match(Month, month.abb),
    Date = as.Date(paste(Year, Month_num, 1, sep = "-"))
  )

plot_df <- ROMS_CUTI_environmental_combined_all_yr_plot %>%
  select(Date,
         DO_avg,
         pH_avg,
         Temperature_avg,
         UpwellWind_avg, 
         CUTI) %>%
  pivot_longer(
    cols = -Date,
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    panel = dplyr::recode(variable,
      DO_avg = "Dissolved Oxygen (mg/L)",
      pH_avg = "pH",
      Temperature_avg = "Temperature (°C)",
      UpwellWind_avg = "Wind (m/s)", 
      CUTI = "Upwelling (m²/s)"
    ),
    panel = factor(panel,
      levels = c("Dissolved Oxygen (mg/L)", "pH", "Temperature (°C)", "Wind (m/s)", "Upwelling (m²/s)")
    )
  )

shade_df <- plot_df %>%
  distinct(year = lubridate::year(Date)) %>%
  mutate(
    xmin = as.Date(paste0(year, "-01-01")),
    xmax = as.Date(paste0(year, "-08-31"))
  )


fig2 <- ggplot(plot_df, aes(x = Date, y = value)) +
  geom_rect(
    data = shade_df,
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
    inherit.aes = FALSE,
    fill = "lightblue",
    alpha = 0.4
  ) +
  geom_line(color = "black", linewidth = 0.7) +
  facet_wrap(~panel, ncol = 1, scales = "free_y", strip.position = "top") +
  scale_x_date(
    date_breaks = "3 years",
    date_labels = "%Y",
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 14) +
  theme(
    strip.text = element_text(size = 16),
    axis.text = element_text(size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
fig2

ggsave("Figures/fig2.png", fig2, width = 7, height = 9, dpi = 300)
```

#### Figure 3 Ocean Basin Indices Time Series

```{r}
# make long data
plot_df <- combined_monthly_indices %>%
  select(Date, mean_sl, mean_pdo, npgo_avg, oni_avg) %>%
  pivot_longer(
    cols = c(npgo_avg, oni_avg, mean_pdo, mean_sl),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    panel = dplyr::recode(
      variable,
      npgo_avg = "NPGO",
      oni_avg  = "ONI",
      mean_pdo = "PDO",
      mean_sl  = "Sea Level Height (m)"
    ),
    panel = factor(panel, levels = c("NPGO", "ONI", "PDO", "Sea Level Height (m)")),
    sign = case_when(
      panel != "Sea Level Height" & value >= 0 ~ "pos",
      panel != "Sea Level Height" & value <  0 ~ "neg",
      TRUE ~ "line"
    )
  )

shade_df <- plot_df %>%
  distinct(year = lubridate::year(Date)) %>%
  mutate(
    xmin = as.Date(paste0(year, "-01-01")),
    xmax = as.Date(paste0(year, "-08-31"))
  )

fig_sup1 <- ggplot() +
  geom_rect(
    data = shade_df,
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
    inherit.aes = FALSE,
    fill = "lightblue",
    alpha = 0.5
  ) +
  # bar panels
  geom_col(
    data = filter(plot_df, panel != "Sea Level Height (m)"),
    aes(Date, value, fill = sign),
    width = 25
  ) +
  # sea level line
  geom_line(
    data = filter(plot_df, panel == "Sea Level Height (m)"),
    aes(Date, value),
    color = "black",
    linewidth = 0.7
  ) +
  facet_wrap(~panel, ncol = 1, scales = "free_y", strip.position = "top") +
  scale_fill_manual(
    values = c(pos = "indianred2", neg = "navy"),
    guide = "none"
  ) +
  scale_x_date(
    date_breaks = "3 years",
    date_labels = "%Y",
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 14) +
  theme(
    strip.text = element_text(size = 16),
    axis.text = element_text(size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

fig_sup1

ggsave("Figures/figsup1.png", fig_sup1, width = 7, height = 9, dpi = 300)
```

## Monthly Correlation Analysis

```{r}
# Join with environmental data
full_data <- left_join(df_all_parameters, goph_recruit, by = "Year")

# List of variables you want to correlate with recruitment
vars_to_correlate <- c("mean_sl", "mean_pdo", "npgo_avg", "oni_avg", "DO_avg", "pH_avg", "Temperature_avg", "CUTI", "UpwellWind_avg")

# Calculate correlations by month
cor_data <- full_data %>%
  pivot_longer(cols = all_of(vars_to_correlate), names_to = "Variable", values_to = "Value") %>%
  group_by(Variable, Month) %>%
  summarise(
    cor_test = list(cor.test(Value, dev, method = "pearson")),
    .groups = "drop"
  ) %>%
  mutate(
    correlation = map_dbl(cor_test, ~ .x$estimate),
    pval        = map_dbl(cor_test, ~ .x$p.value),
    t_stat      = map_dbl(cor_test, ~ .x$statistic),
    significant = pval < 0.05
  ) %>%
  select(-cor_test)



# Order months properly
cor_data$Month <- factor(cor_data$Month, levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

cor_data <- cor_data %>%
  mutate(Variable = dplyr::recode(as.character(Variable),
    "mean_pdo"        = "PDO",
    "mean_sl"         = "Sea Level",
    "npgo_avg"        = "NPGO",
    "oni_avg"         = "ONI",
    "DO_avg" = "DO",
    "pH_avg"  = "pH",
    "Temperature_avg" = "Temperature",
    "UpwellWind_avg" = "Wind", 
    "CUTI" = "Upwelling"
  ))
```

### [REV #1 & #2] Multiple-comparison correction and autocorrelation-adjusted significance

The block above computes ~96 Pearson tests (9 variables x 12 months, minus any
all-NA cells) and flags raw p < 0.05. Reviewers asked us to (a) correct for
multiple testing and (b) account for temporal autocorrelation, which inflates
the apparent significance of correlations between two autocorrelated annual
series (recruitment deviations and, especially, decadal indices like PDO).

We do both here without altering the original `cor_data` object, so every
downstream chunk that uses `cor_data` still runs unchanged. The augmented
object is `cor_data_adj`.

```{r rev12-setup}
# ---- Effective sample size (Pyper & Peterman 1998, modified Chelton) ----
# For two series X and Y of length N, the effective number of independent
# observations is:
#   1/Neff = 1/N + (2/N) * sum_{j=1}^{L} ((N-j)/N) * rho_X(j) * rho_Y(j)
# where rho_.(j) are the lag-j autocorrelations. Following Pyper & Peterman,
# the sum is truncated at L = floor(N/5) to reduce estimation bias, and Neff
# is capped at N.
effective_n <- function(x, y, lag_prop = 1/5) {
  ok <- stats::complete.cases(x, y)
  x  <- x[ok]; y <- y[ok]
  n  <- length(x)
  if (n < 5 || stats::sd(x) == 0 || stats::sd(y) == 0) return(as.numeric(n))
  L  <- max(1L, floor(n * lag_prop))
  # acf()[-1] drops the lag-0 term (which is 1); demean = TRUE by default
  ax <- stats::acf(x, lag.max = L, plot = FALSE, demean = TRUE)$acf[-1]
  ay <- stats::acf(y, lag.max = L, plot = FALSE, demean = TRUE)$acf[-1]
  j  <- seq_len(L)
  inv_neff <- 1/n + (2/n) * sum(((n - j)/n) * ax * ay)
  neff <- 1 / inv_neff
  # keep Neff in a sensible range: at least 3 (so df >= 1), at most n
  min(max(neff, 3), n)
}

# p-value from a correlation r given an (effective) sample size
p_from_r <- function(r, n_eff) {
  df <- n_eff - 2
  if (df <= 0 || is.na(r) || abs(r) >= 1) return(NA_real_)
  t  <- r * sqrt(df / (1 - r^2))
  2 * stats::pt(-abs(t), df)
}
```

```{r rev12-compute}
# Recompute correlations while retaining the per-year vectors so we can
# estimate autocorrelation. Ordered by Year within each Variable x Month.
cor_data_adj <- full_data %>%
  dplyr::select(Year, Month, dplyr_dev = dev, dplyr::all_of(vars_to_correlate)) %>%
  tidyr::pivot_longer(cols = dplyr::all_of(vars_to_correlate),
                      names_to = "Variable", values_to = "Value") %>%
  dplyr::rename(dev = dplyr_dev) %>%
  dplyr::arrange(Variable, Month, Year) %>%
  dplyr::group_by(Variable, Month) %>%
  dplyr::summarise(
    n_obs   = sum(stats::complete.cases(Value, dev)),
    r       = tryCatch(suppressWarnings(stats::cor(Value, dev, use = "complete.obs")),
                       error = function(e) NA_real_),
    p_raw   = tryCatch(stats::cor.test(Value, dev)$p.value, error = function(e) NA_real_),
    n_eff   = effective_n(Value, dev),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    p_autocorr = mapply(p_from_r, r, n_eff)
  )

# ---- Comment #1: Benjamini-Hochberg FDR across the whole screen ----
# Applied to the raw p-values, and (Comment #1 + #2 combined) to the
# autocorrelation-adjusted p-values.
cor_data_adj <- cor_data_adj %>%
  dplyr::mutate(
    # n set to the number of tests actually performed (non-NA cells) so the
    # FDR denominator is exact even if a Variable x Month cell was all-NA.
    p_raw_fdr       = stats::p.adjust(p_raw,      method = "BH", n = sum(!is.na(p_raw))),
    p_autocorr_fdr  = stats::p.adjust(p_autocorr, method = "BH", n = sum(!is.na(p_autocorr))),
    sig_raw         = p_raw          < 0.05,
    sig_fdr         = p_raw_fdr      < 0.05,   # survives FDR
    sig_autocorr    = p_autocorr     < 0.05,   # survives autocorrelation adj.
    sig_both        = p_autocorr_fdr < 0.05    # survives FDR + autocorrelation
  )

# Nice variable labels (same mapping used for cor_data) + month ordering
recode_vars <- c(
  "mean_pdo" = "PDO", "mean_sl" = "Sea Level", "npgo_avg" = "NPGO",
  "oni_avg" = "ONI", "DO_avg" = "DO", "pH_avg" = "pH",
  "Temperature_avg" = "Temperature", "UpwellWind_avg" = "Wind", "CUTI" = "Upwelling"
)
cor_data_adj <- cor_data_adj %>%
  dplyr::mutate(
    Variable = dplyr::recode(as.character(Variable), !!!recode_vars),
    Month    = factor(Month, levels = month.abb)
  )

# How many tests were actually run (non-NA), and how many survive each filter?
rev12_counts <- cor_data_adj %>%
  dplyr::summarise(
    n_tests            = sum(!is.na(p_raw)),
    n_sig_raw          = sum(sig_raw,      na.rm = TRUE),
    n_sig_fdr          = sum(sig_fdr,      na.rm = TRUE),
    n_sig_autocorr     = sum(sig_autocorr, na.rm = TRUE),
    n_sig_fdr_autocorr = sum(sig_both,     na.rm = TRUE)
  )
rev12_counts
```

```{r rev12-blocks}
# ---- Comment #1: flag CONTIGUOUS blocks of raw-significant months ----
# Reviewers noted that blocks of adjacent significant months (e.g. pH Jan-Apr;
# Temperature Feb-Mar) are more credible than isolated single-month hits.
# We label, for each variable, runs of consecutive months with raw p < 0.05
# and mark those of length >= 2 as "blocks".
cor_data_adj <- cor_data_adj %>%
  dplyr::arrange(Variable, Month) %>%
  dplyr::group_by(Variable) %>%
  dplyr::mutate(
    .sig = tidyr::replace_na(sig_raw, FALSE),
    run_id = cumsum(.sig != dplyr::lag(.sig, default = FALSE)),
  ) %>%
  dplyr::group_by(Variable, run_id) %>%
  dplyr::mutate(
    block_len   = sum(.sig),
    in_block    = .sig & block_len >= 2
  ) %>%
  dplyr::ungroup() %>%
  dplyr::select(-.sig, -run_id)

# Human-readable summary of contiguous significant blocks
significant_blocks <- cor_data_adj %>%
  dplyr::filter(in_block) %>%
  dplyr::group_by(Variable, block_len) %>%
  dplyr::summarise(months = paste(as.character(Month), collapse = ", "),
                   .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(block_len))
significant_blocks

# Full augmented results table (rounded), useful for a supplementary table
rev12_table <- cor_data_adj %>%
  dplyr::transmute(
    Variable, Month,
    n = n_obs,
    r = round(r, 3),
    `N_eff` = round(n_eff, 1),
    `p (raw)`            = signif(p_raw, 3),
    `p (FDR)`            = signif(p_raw_fdr, 3),
    `p (autocorr)`       = signif(p_autocorr, 3),
    `p (FDR+autocorr)`   = signif(p_autocorr_fdr, 3),
    in_block
  ) %>%
  dplyr::arrange(Variable, Month)
knitr::kable(rev12_table)
```

#### Figure 4 - Correlation plot

```{r}

desired_order <- c("DO","pH","Temperature", "Wind", "Upwelling", "PDO","Sea Level", "NPGO", "ONI")

cor_data$Variable <- factor(cor_data$Variable, levels = desired_order)

fig4 <- ggplot(cor_data, aes(x = Month, y = correlation, color = significant, group = 1)) +
  geom_point(size = 3) +
 geom_line(size = 1) +
  facet_wrap(~Variable, ncol = 2) +
  geom_hline(yintercept = 0, linetype = "solid", size = 0.25) +
  geom_hline(yintercept = c(-0.25, 0.25), linetype = "dashed", color = "black", size = 0.75) +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "indianred")) +
  theme_classic() +
  labs(
    title = "",
    y = "Correlation with Recruitment",
    color = "Significant"
  )+   theme_bw() +
  theme(strip.text = element_text(size = 12, face = "bold"), 
        axis.line = element_line(color = "black"),
        legend.position = "none",
        axis.title = element_text(size = 16),   # Axis titles
        axis.text = element_text(size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1),  # 
        title = element_text(size = 20))
fig4

# Making a data frame with significant months
significant_months <- cor_data %>%
  filter(significant == TRUE)

ggsave("Figures/fig4.png", fig4, width = 12, height = 12, dpi = 300)
```

### [REV #1 & #2] Figure 4 — significance tiers after correction

Same panel layout as the original Figure 4, but each month is now shaded by how
robust its correlation is: raw p < 0.05 only, surviving FDR, or surviving both
FDR and the autocorrelation adjustment. Points inside a contiguous run of
raw-significant months (length >= 2) are outlined, since reviewers view those
blocks as more credible than isolated single-month hits.

```{r rev12-fig4}
cor_data_adj_plot <- cor_data_adj %>%
  dplyr::mutate(
    Variable = factor(Variable, levels = desired_order),
    tier = dplyr::case_when(
      sig_both     ~ "FDR + autocorrelation",
      sig_fdr      ~ "Survives FDR",
      sig_raw      ~ "Raw p < 0.05 only",
      TRUE         ~ "n.s."
    ),
    tier = factor(tier, levels = c("n.s.", "Raw p < 0.05 only",
                                   "Survives FDR", "FDR + autocorrelation"))
  )

tier_colors <- c(
  "n.s."                  = "grey70",
  "Raw p < 0.05 only"     = "#F4A261",
  "Survives FDR"          = "#E76F51",
  "FDR + autocorrelation" = "#9B2226"
)

fig4_rev <- ggplot(cor_data_adj_plot,
                   aes(x = Month, y = r, group = 1)) +
  geom_hline(yintercept = 0, linetype = "solid", size = 0.25) +
  geom_hline(yintercept = c(-0.25, 0.25), linetype = "dashed",
             color = "black", size = 0.5) +
  geom_line(size = 0.8, color = "grey40") +
  # black ring on points that fall inside a contiguous significant block
  geom_point(data = dplyr::filter(cor_data_adj_plot, in_block),
             aes(x = Month, y = r), shape = 21, size = 5,
             fill = NA, color = "black", stroke = 1.1) +
  geom_point(aes(color = tier), size = 3) +
  facet_wrap(~Variable, ncol = 2) +
  scale_color_manual(values = tier_colors, name = NULL, drop = FALSE) +
  labs(y = "Correlation with Recruitment", x = NULL) +
  theme_bw() +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.line = element_line(color = "black"),
    legend.position = "bottom",
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
fig4_rev

ggsave("Figures/fig4_revised.png", fig4_rev, width = 12, height = 13, dpi = 300)
```

## PSLR Analysis

### PSLR averaging from Jan to Aug

```{r}
# Average oceanographic parameters annually 
df_all_parameters_feb_sept_average <- df_all_parameters %>% 
  filter(Month %in% c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug")) %>% 
  group_by(Year) %>%
  summarise(across(
    c(DO_avg, pH_avg, Temperature_avg, UpwellWind_avg,CUTI, mean_pdo, npgo_avg, mean_sl, oni_avg),
    ~ mean(.x, na.rm = TRUE),
    .names = "mean_{.col}"
  )) %>%
  ungroup()

# combined 
df_gopher_avg_mar_sept <-left_join(df_all_parameters_feb_sept_average,goph_recruit, by = "Year")

df_gopher_avg_mar_sept <- df_gopher_avg_mar_sept %>% 
  select(dev, mean_mean_pdo, mean_CUTI, mean_mean_sl,mean_pH_avg, mean_DO_avg, mean_Temperature_avg) %>% 
  rename(deviation = dev, 
         CUTI = mean_CUTI,
         PDO = mean_mean_pdo,
         "Sea level height" = mean_mean_sl, 
         pH = mean_pH_avg, 
         DO = mean_DO_avg, 
         Temperature = mean_Temperature_avg)

# Define X and Y together and filter complete cases
X_raw <- df_gopher_avg_mar_sept %>% 
  select(PDO, CUTI,  "Sea level height" ,pH, DO, Temperature)

# Combine into one frame with response
data_complete <- cbind(dev = df_gopher_avg_mar_sept$deviation, X_raw) %>% 
  as.data.frame() %>% 
  na.omit()  # remove rows with any NA

# Now separate into X and Y
Y <- data_complete$dev
X <- as.matrix(data_complete %>% select(-dev))

plsr_model <- plsr(Y ~ X, scale = TRUE, validation = "LOO")
pls_scores <- scores(plsr_model)

df_pls <- data.frame(
  dev = Y,
  Comp1 = pls_scores[, 1],
  Comp2 = pls_scores[, 2]
)

# Summary of model
summary(plsr_model)

# Root Mean Squared Error of Prediction
RMSEP(plsr_model)

# Choose number of components based on RMSEP or % variance explained
plot(RMSEP(plsr_model), legendpos = "topright")

# Variance explained
explvar(plsr_model)  # % of variance explained in X by each component

# Component loadings: how original vars contribute to latent variables
loadings(plsr_model)

# Scores: the new coordinate representation of observations
pls_scores <- scores(plsr_model)

# Biplot: visualize scores and loadings
biplot(plsr_model, comps = 1:2)

# Fit a linear model using first 2 PLS components
lm_pls <- lm(dev ~ Comp1 + Comp2, data = df_pls)
summary(lm_pls)

# check residuals 
simulated_res <- simulateResiduals(fittedModel = lm_pls)
plot(simulated_res) #these look good

selectNcomp(plsr_model, method = "onesigma")

lm_pls_1comp <- lm(dev ~ Comp1, data = df_pls)
summary(lm_pls_1comp)

plot(RMSEP(plsr_model), legendpos = "topright")

# 1-component model is optimal, so we use opt.comp = 1
vip_scores <- VIP(plsr_model, opt.comp = 1)

# Turn into data frame properly
vip_df <- data.frame(
  Variable = names(vip_scores),
  VIP = vip_scores
)

ggplot(vip_df, aes(x = reorder(Variable, VIP), y = VIP)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "indianred4", size = 1) +
  geom_hline(yintercept = .8, linetype = "dashed", color = "indianred3", size = 1) +
  coord_flip() +
  theme_classic() +
  labs(title = "Variable Importance in Projection (VIP)",
       x = "Predictor",
       y = "VIP Score")
```

#### Combined figure for PSLR

```{r}
# Extract predictor names from X
predictor_names <- colnames(X)

# ========== FIGURE A: PLS BIPLOT ==========

loadings_data <- loadings(plsr_model)[, 1:2]
loadings_df <- data.frame(
  Variable = rownames(loadings_data),
  Component1 = loadings_data[, 1],
  Component2 = loadings_data[, 2]
)

fig_a_repro <- ggplot(loadings_df, aes(x = Component1, y = Component2)) +
  # Unit circle
  geom_path(data = data.frame(
    x = cos(seq(0, 2 * pi, length.out = 100)),
    y = sin(seq(0, 2 * pi, length.out = 100))
  ), aes(x = x, y = y), color = "black", linewidth = 1) +

  # Arrows and labels
  geom_segment(aes(x = 0, y = 0, xend = Component1, yend = Component2),
               arrow = arrow(length = unit(0.3, "cm")), 
               color = "steelblue", linewidth = 1.2) +
  geom_text(aes(label = Variable), 
            color = "steelblue", size = 4, fontface = "bold",
            nudge_x = ifelse(loadings_df$Component1 > 0, 0.05, -0.05),
            nudge_y = ifelse(loadings_df$Component2 > 0, 0.05, -0.05)) +
  
  # Optional: Response direction (illustrative only)
  # You can calculate actual correlation direction if desired
  geom_segment(aes(x = 0, y = 0, xend = 0.6, yend = 0.4), 
               arrow = arrow(length = unit(0.5, "cm")), 
               color = "indianred", linewidth = 1.5) +
  geom_text(aes(x = 0.68, y = 0.45, label = "Recruitment"), 
             color = "indianred", size = 4, fontface = "bold") +
  
  xlim(-1.1, 1.1) + ylim(-1.1, 1.1) +
  labs(x = "Component 1", y = "Component 2", title = "Reproductive Season Average") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  ) +
  coord_fixed()
# ========== FIGURE B: COMPONENT 1 vs RESPONSE ==========

# Ensure response is matched to filtered data
component1_scores <- scores(plsr_model)[, 1]
response_var <- Y  # this Y comes from the filtered, complete-case data

# Compute R² and p-value
var_explained <- explvar(plsr_model)[1]
lm_comp1 <- lm(response_var ~ component1_scores)
lm_summary <- summary(lm_comp1)
p_value <- lm_summary$coefficients[2, 4]
r_squared <- lm_summary$r.squared

# Format p-value
p_text <- if (p_value < 0.001) {
  "p < 0.001"
} else if (p_value < 0.01) {
  paste0("p = ", format(round(p_value, 3), nsmall = 3))
} else {
  paste0("p = ", format(round(p_value, 2), nsmall = 2))
}

annotation_text <- paste0("R² = ", format(round(r_squared, 3), nsmall = 3), "\n", p_text)

fig_b_repro <- ggplot(data.frame(Component1 = component1_scores, Response = response_var),
                aes(x = Component1, y = Response)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_smooth(method = "lm", color = "steelblue", fill = "black", alpha = 0.3) +
  annotate("text", 
           x = Inf, y = Inf, 
           label = annotation_text,
           hjust = 1.3, vjust = 1.3,
           size = 4, 
           fontface = "bold",
           color = "black") +
  labs(
    x = paste0("PLSR component 1 (", round(var_explained, 1), "%)"),
    y = "Recruitment Deviations",
    title = ""
  ) +
  theme_classic() +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )


# ========== FIGURE C: VIP PLOT ==========

fig_c_repro <- ggplot(vip_df, aes(x = reorder(Variable, VIP), y = VIP)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "indianred4", size = 1) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "indianred3", size = 1) +
  coord_flip() +
  theme_classic() +
  labs(title = "",
       x = "Predictor",
       y = "VIP Score") +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )


# ========== COMBINE ALL THREE FIGURES ==========
combined_plot <- fig_a_repro + fig_b_repro + fig_c_repro + 
  plot_layout(ncol = 3)
combined_plot

ggsave('Figures/fig5_repro.png', width = 10, height = 4, dpi = 300)
```

## Best historical predictor - including CUTI just for fun

```{r}
## Join everything
df_env_joined <- left_join(ROMS_CUTI_environmental_combined_all_yr, goph_recruit, by = "Year") %>%
  select(Year, Month, DO_avg, pH_avg, Temperature_avg, dev, CUTI) %>% 
  rename('CUTI_avg' = "CUTI")

## Puts drivers into long format and calculates rolling averages 
### Calculates trailing rolling averages - so the rolling average ends in the month it states
### EX - 3-month in "Jul" is May, June, July
month_levels <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

df_long <- df_env_joined %>%
  pivot_longer(
    cols = c(DO_avg, pH_avg, Temperature_avg, CUTI_avg),
    names_to = "driver",
    values_to = "value"
  ) %>%
  mutate(
    Month = factor(Month, levels = month_levels, ordered = TRUE),
    Month_num = match(as.character(Month), month_levels)
  ) %>%
  arrange(driver, Year, Month_num)

df_windows <- df_long %>%
  group_by(driver, Year) %>%
  arrange(Month_num, .by_group = TRUE) %>%
  mutate(
    value_1mo = value,
    value_3mo = slide_dbl(value, mean, .before = 2, .complete = TRUE, na.rm = FALSE), 
    value_6mo = slide_dbl(value, mean, .before = 6, .complete = TRUE, na.rm = FALSE)
  ) %>%
  ungroup()



df_candidates <- df_windows %>%
  select(Year, Month, dev, driver, value_1mo, value_3mo, value_6mo) %>%
  pivot_longer(
    cols = starts_with("value_"),
    names_to = "window",
    values_to = "predictor"
  ) %>%
  mutate(
    window = dplyr::recode(
      window,
      value_1mo = "single month",
      value_3mo = "3-month rolling avg",
      value_6mo = "6-month rolling avg"
    )
  ) %>%
  filter(!is.na(predictor), !is.na(dev))

## Fit every model and choose best per driver
model_results <- df_candidates %>%
  group_by(driver, Month, window) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(dev ~ predictor, data = .x)),
    glance = map(model, glance),
    tidy = map(model, tidy)
  ) %>%
  mutate(
    r.squared = map_dbl(glance, "r.squared"),
    adj.r.squared = map_dbl(glance, "adj.r.squared"),
    model_p = map_dbl(glance, "p.value"),
    slope = map_dbl(tidy, ~ .x %>% filter(term == "predictor") %>% pull(estimate)),
    slope_p = map_dbl(tidy, ~ .x %>% filter(term == "predictor") %>% pull(p.value))
  ) %>%
  select(driver, Month, window, model, r.squared, adj.r.squared, model_p, slope, slope_p)

best_models <- model_results %>%
  group_by(driver) %>%
  slice_max(order_by = r.squared, n = 1, with_ties = FALSE) %>%
  ungroup()

best_models

## Get fitted values and confidence intervals for the best model of each driver
best_predictions <- best_models %>%
  select(driver, Month, window, model, r.squared) %>%
  left_join(df_candidates, by = c("driver", "Month", "window")) %>%
  group_by(driver, Month, window, r.squared) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(dev ~ predictor, data = .x)),
    pred = map2(model, data, ~ {
      preds <- predict(.x, newdata = .y, interval = "confidence")
      bind_cols(.y, as_tibble(preds))
    })
  ) %>%
  select(-data, -model) %>%
  unnest(pred) %>%
  ungroup()
```

### [REV #3] Reporting the full window search (model-selection inflation)

The single-driver models above search single-month, 3-month, and 6-month
windows across all months and keep the best-performing window per driver
(`best_models`, via `slice_max` on r-squared). Selecting the strongest window
from many candidates inflates the reported r-squared and understates
uncertainty. As reviewers requested, we now (a) state how many candidate
windows were searched per driver, (b) show the *full distribution* of
in-sample r-squared across all of those candidates so the selected value can be
read against the field it was chosen from, and (c) report a leave-one-out (LOO)
*predictive* R-squared for each model, which — unlike in-sample r-squared —
does not reward overfitting to a particular window.

```{r rev3-loo-fn}
# Leave-one-out predictive R^2 = 1 - PRESS/TSS.
# Can be negative when the model predicts worse than the series mean.
loo_r2 <- function(mod) {
  d <- mod$model                 # columns: dev, predictor
  n <- nrow(d)
  if (n < 3L) return(NA_real_)
  err <- vapply(seq_len(n), function(i) {
    fit_i <- stats::lm(dev ~ predictor, data = d[-i, , drop = FALSE])
    d$dev[i] - stats::predict(fit_i, newdata = d[i, , drop = FALSE])
  }, numeric(1))
  press <- sum(err^2)
  tss   <- sum((d$dev - mean(d$dev))^2)
  1 - press / tss
}
```

```{r rev3-distribution}
# In-sample r^2 AND out-of-sample (LOO) R^2 for every candidate window
window_distribution <- model_results %>%
  dplyr::ungroup() %>%
  dplyr::mutate(loo_r2 = purrr::map_dbl(model, loo_r2))

# (a) How many candidate windows were searched per driver
candidate_counts <- window_distribution %>%
  dplyr::count(driver, name = "n_candidate_windows")
candidate_counts

# Distribution of in-sample r^2 within each driver's candidate set
r2_distribution_summary <- window_distribution %>%
  dplyr::group_by(driver) %>%
  dplyr::summarise(
    n_windows   = dplyr::n(),
    r2_min      = round(min(r.squared, na.rm = TRUE), 3),
    r2_median   = round(stats::median(r.squared, na.rm = TRUE), 3),
    r2_max      = round(max(r.squared, na.rm = TRUE), 3),
    .groups = "drop"
  )
r2_distribution_summary

# (c) The SELECTED best window, with in-sample vs LOO predictive R^2.
# `shrinkage` = how much apparent skill is lost out-of-sample; a large value
# is the signature of window-selection overfitting.
best_with_loo <- window_distribution %>%
  dplyr::group_by(driver) %>%
  dplyr::slice_max(r.squared, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::left_join(candidate_counts, by = "driver") %>%
  dplyr::transmute(
    Driver = dplyr::recode(driver,
      DO_avg = "Dissolved Oxygen", pH_avg = "pH",
      Temperature_avg = "Temperature", CUTI_avg = "CUTI"),
    `Selected window`      = window,
    `Candidates searched`  = n_candidate_windows,
    `r² (in-sample)`       = round(r.squared, 3),
    `R² (LOO predictive)`  = round(loo_r2, 3),
    `Shrinkage`            = round(r.squared - loo_r2, 3)
  )
knitr::kable(best_with_loo)
```

```{r rev3-figure}
# (b) Full distribution of candidate r^2 per driver, with the selected
# window highlighted. Makes the "forking paths" search visible to readers.
best_points <- window_distribution %>%
  dplyr::group_by(driver) %>%
  dplyr::slice_max(r.squared, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup()

driver_lab <- c(DO_avg = "Dissolved Oxygen", pH_avg = "pH",
                Temperature_avg = "Temperature", CUTI_avg = "CUTI")

fig_window_search <- ggplot(window_distribution,
                            aes(x = r.squared,
                                y = dplyr::recode(driver, !!!driver_lab))) +
  geom_jitter(height = 0.12, width = 0, alpha = 0.35, size = 1.6,
              color = "grey40") +
  geom_point(data = best_points,
             aes(x = r.squared,
                 y = dplyr::recode(driver, !!!driver_lab)),
             color = "#9B2226", size = 3.5) +
  labs(
    x = expression("Candidate-window r"^2),
    y = NULL
  ) +
  theme_bw(base_size = 14) +
  theme(panel.grid.minor = element_blank())
fig_window_search

ggsave("Figures/fig_window_search.png", fig_window_search,
       width = 7, height = 4, dpi = 300)
```

#### Table with linear model outputs

```{r}
month_abbs <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

make_window_label <- function(month, window_type) {
  month <- as.integer(month)
  
  if (window_type == "single month") {
    return(month_abbs[month])
  }
  
  if (window_type == "3-month rolling avg") {
    return(paste0(month_abbs[month - 2], "-", month_abbs[month]))
  }
  
  if (window_type == "6-month rolling avg") {
    return(paste0(month_abbs[month - 5], "-", month_abbs[month]))
  }
  
  return(NA_character_)
}

driver_labels <- c(
  DO_avg = "Dissolved Oxygen",
  pH_avg = "pH",
  Temperature_avg = "Temperature", 
  CUTI_avg = "CUTI"
)

best_model_table <- best_models %>%
  mutate(
    Month_num = as.integer(Month),
    driver_name = dplyr::recode(driver, !!!driver_labels),
    period = map2_chr(Month_num, window, make_window_label),
    window_label = case_when(
      window == "single month" ~ period,
      TRUE ~ paste(period, "mean")
    ),
    tidy_out = map(model, tidy),
    glance_out = map(model, glance)
  ) %>%
  mutate(
    intercept = map_dbl(tidy_out, ~ .x %>% filter(term == "(Intercept)") %>% pull(estimate)),
    slope = map_dbl(tidy_out, ~ .x %>% filter(term == "predictor") %>% pull(estimate)),
    slope_se = map_dbl(tidy_out, ~ .x %>% filter(term == "predictor") %>% pull(std.error)),
    slope_t = map_dbl(tidy_out, ~ .x %>% filter(term == "predictor") %>% pull(statistic)),
    slope_p = map_dbl(tidy_out, ~ .x %>% filter(term == "predictor") %>% pull(p.value)),
    r2 = map_dbl(glance_out, "r.squared"),
    adj_r2 = map_dbl(glance_out, "adj.r.squared"),
    model_p = map_dbl(glance_out, "p.value"),
    AIC = map_dbl(glance_out, "AIC"),
    BIC = map_dbl(glance_out, "BIC"),
    n = map_dbl(glance_out, "nobs")
  ) %>%
  transmute(
    Driver = driver_name,
    `Selected window` = window_label,
    Intercept = round(intercept, 3),
    Slope = round(slope, 3),
    `SE` = round(slope_se, 3),
    `t value` = round(slope_t, 3),
    `Slope p` = signif(slope_p, 3),
    `R²` = round(r2, 3),
    `Adj. R²` = round(adj_r2, 3),
    `p-value` = signif(model_p, 3),
    AIC = round(AIC, 2),
    BIC = round(BIC, 2),
    n = as.integer(n)
  )

best_model_table


best_model_table1 <- best_model_table %>% 
  select(Driver, `Selected window`, n, Slope, `SE`, `R²`, `p-value`)
knitr::kable(
  best_model_table1,
  #caption = "Summary statistics for the strongest univariate linear models relating annual recruitment deviation to each environmental driver."
)
```

#### Figure 6. Linear regression predictions

```{r}
observed_dev <- df_env_joined %>%
  distinct(Year, dev)

driver_labels <- c(
  DO_avg = "Dissolved Oxygen",
  pH_avg = "pH",
  Temperature_avg = "Temperature", 
  CUTI_avg = "CUTI"
)

driver_colors <- c(
  Temperature_avg = "#E76F51",
  DO_avg = "#4C78A8",
  pH_avg = "#66A61E", 
  CUTI_avg = "chocolate4"
)

month_labels <- c(
  "1" = "Jan", "2" = "Feb", "3" = "Mar", "4" = "Apr",
  "5" = "May", "6" = "Jun", "7" = "Jul", "8" = "Aug",
  "9" = "Sep", "10" = "Oct", "11" = "Nov", "12" = "Dec"
)

subtitle_df <- best_models %>%
  mutate(
    driver_name = dplyr::recode(driver, !!!driver_labels),
    month_name = dplyr::recode(as.character(Month), !!!month_labels),
    label = glue::glue("{driver_name}: {month_name}, {window}, r²={round(r.squared, 3)}")
  )

subtitle_df

label_positions <- best_predictions %>%
  group_by(driver) %>%
  filter(Year == max(Year)) %>%
  distinct(driver, Year, fit) %>%
  left_join(
    best_models %>% select(driver, r.squared),
    by = "driver"
  ) %>%
  mutate(
    label = paste0("r² = ", round(r.squared, 2))
  )

label_positions <- label_positions %>%
  arrange(driver) %>%
  mutate(
    y_offset = case_when(
      driver == "DO_avg" ~ 0.25,
      driver == "pH_avg" ~ -0.0,
      driver == "Temperature_avg" ~ 0.12,
      driver == "CUTI_avg" ~ -.1
    )
  )

subtitle_text <- paste(subtitle_df$label, collapse = "\n")

ggplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  
  geom_ribbon(
    data = best_predictions,
    aes(x = Year, ymin = lwr, ymax = upr, fill = driver),
    alpha = 0.2
  ) +
  
  geom_line(
    data = best_predictions,
    aes(x = Year, y = fit, color = driver),
    linetype = "dashed",
    linewidth = 1
  ) +

  geom_line(
    data = observed_dev,
    aes(x = Year, y = dev),
    color = "black",
    linewidth = 1
  ) +
  
  geom_text(
    data = label_positions,
    aes(x = Year, y = fit + y_offset, label = label, color = driver),
    hjust = -0.1,
    size = 4,
    show.legend = FALSE
  ) +
  
  scale_color_manual(
    values = driver_colors,
    labels = driver_labels,
    name = NULL
  ) +
  scale_fill_manual(
    values = driver_colors,
    labels = driver_labels,
    name = NULL
  ) +
  
  labs(
    title = "",
    x = "Year",
    y = "Recruitment deviation"
  ) +
  
  coord_cartesian(
    xlim = c(min(observed_dev$Year), max(observed_dev$Year) + 1)
  ) +
  
  theme_bw(base_size = 16) +
  theme(
    legend.position = "bottom"
  )

ggsave('Figures/fig6.png', width = 8, height = 4, dpi = 300)
```

#### Better at predicting negative vs. positive recruitment
```{r}
sign_performance <- best_models %>%
  mutate(
    augmented = map(model, augment)
  ) %>%
  unnest(augmented) %>%
  mutate(
    observed_sign = if_else(dev > 0, "positive", "negative"),
    predicted_sign = if_else(.fitted > 0, "positive", "negative"),
    correct = observed_sign == predicted_sign
  )

accuracy_summary <- sign_performance %>%
  group_by(driver) %>%
  summarise(
    accuracy = mean(correct),
    n = n()
  )

loocv_classification <- function(model) {
  data <- model$model
  n <- nrow(data)
  
  preds <- map_dbl(1:n, function(i) {
    train <- data[-i, ]
    test <- data[i, ]
    
    m <- lm(dev ~ predictor, data = train)
    predict(m, newdata = test)
  })
  
  tibble(
    observed = data$dev,
    predicted = preds
  ) %>%
    mutate(
      observed_sign = if_else(observed > 0, "positive", "negative"),
      predicted_sign = if_else(predicted > 0, "positive", "negative"),
      correct = observed_sign == predicted_sign
    )
}

loocv_results <- best_models %>%
  mutate(
    results = map(model, loocv_classification)
  ) %>%
  unnest(results)

loocv_summary <- loocv_results %>%
  group_by(driver) %>%
  summarise(
    accuracy = mean(correct),
    TP = sum(predicted_sign == "positive" & observed_sign == "positive"),
    TN = sum(predicted_sign == "negative" & observed_sign == "negative"),
    FP = sum(predicted_sign == "positive" & observed_sign == "negative"),
    FN = sum(predicted_sign == "negative" & observed_sign == "positive"),
    sensitivity = TP / (TP + FN),
    specificity = TN / (TN + FP)
  )

loocv_summary

classification_table <- loocv_results %>%
  group_by(driver) %>%
  summarise(
    n = n(),
    TP = sum(predicted_sign == "positive" & observed_sign == "positive"),
    TN = sum(predicted_sign == "negative" & observed_sign == "negative"),
    FP = sum(predicted_sign == "positive" & observed_sign == "negative"),
    FN = sum(predicted_sign == "negative" & observed_sign == "positive"),
    
    Accuracy = (TP + TN) / n,
    Sensitivity = TP / (TP + FN),   # detects positive years
    Specificity = TN / (TN + FP)    # detects negative years
  ) %>%
  mutate(
    Driver = dplyr::recode(driver,
      DO_avg = "Dissolved Oxygen",
      pH_avg = "pH",
      Temperature_avg = "Temperature"
    )
  ) %>%
  select(Driver, n, Accuracy, Sensitivity, Specificity, TP, TN, FP, FN)

classification_table

final_table <- classification_table %>%
  mutate(
    Accuracy = round(Accuracy, 2),
    Sensitivity = round(Sensitivity, 2),
    Specificity = round(Specificity, 2)
  ) %>%
  select(Driver, Accuracy, Sensitivity, Specificity)

final_table

knitr::kable(
  final_table
)

```

#### Saving as a .rds for future projections

```{r}
best_model_objects <- best_models %>%
  select(driver, Month, window, model)

saveRDS(best_model_objects, "best_historical_recruitment_models.rds")
```

## Future Oceanographic Characteristics

```{r}
full_palette <- wes_palette("Moonrise2", type = "discrete")
ESM_colors <- c("GFDL" = full_palette[1], 
                      "HADL" = full_palette[2],
                 "IPSL" = full_palette[4])
```

#### Reading in csv that were cleaned in python
```{r}
GFDL<- read.csv("Data/GFDL_monthly_averages_allyear_2000_2100_ALL_41to34p4.csv") %>%
  mutate(ESM = "GFDL")

HADL <- read.csv("Data/HADL_monthly_averages_allyear_2000_2100_ALL_41to34p4.csv") %>%
  mutate(ESM = "HADL")

IPSL <- read.csv("Data/IPSL_monthly_averages_allyear_2000_2100_ALL_41to34p4.csv") %>%
  mutate(ESM = "IPSL")

# Combine all into one dataframe
future_ocean <- bind_rows(
  GFDL, HADL,IPSL
)

# View result
head(future_ocean)
```

#### read in future upwelling index from Mike's Paper 
```{r}
future_upwelling <- read.csv("Data/future_cuti_all_models.csv") %>% 
  filter(year > 1999) %>% 
  filter(between(latitude, 35, 41)) %>% 
  group_by(model, year, month) %>%
  summarise(upwelling_avg_lat = mean(upwelling_index, na.rm = TRUE),
            .groups = "drop") %>% 
  rename(
    Year = year,
    Month = month,
    ESM = model,
    CUTI = upwelling_avg_lat
  ) %>% 
  mutate(Month = month.abb[Month])%>%
  mutate(ESM = toupper(ESM)) #makes it caps

future_upwelling_avg <- future_upwelling %>%
  filter(Month >= 7 & Month <= 12) %>%
  group_by(model, year) %>%
  summarise(upwelling_avg = mean(upwelling_index, na.rm = TRUE),
            .groups = "drop")
  
```

#### Cleaning data for plotting
```{r}
df_joined <- future_ocean %>%
  left_join(future_upwelling, by = c("Year", "Month", "ESM"))

future_ocean2 <- df_joined %>% 
  filter(Month %in% c("Jan", 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug'))
# annual means by ESM

future_annual <- future_ocean2 %>%
  group_by(ESM, Year) %>%
  summarize(
    DO = mean(DO_avg, na.rm = TRUE),
    pH = mean(pH_avg, na.rm = TRUE),
    Temperature = mean(Temperature_avg, na.rm = TRUE),
    Upwelling = mean(CUTI, na.rm = TRUE),
    .groups = "drop"
  )

# long format
future_long <- future_annual %>%
  pivot_longer(
    cols = c(DO, pH, Temperature),
    names_to = "Variable",
    values_to = "Value"
  )

# 1. Annual means by ESM
future_annual <- future_ocean2 %>%
  group_by(ESM, Year) %>%
  summarize(
    DO = mean(DO_avg, na.rm = TRUE),
    pH = mean(pH_avg, na.rm = TRUE),
    Temperature = mean(Temperature_avg, na.rm = TRUE),
    Upwelling = mean(CUTI, na.rm = TRUE),
    .groups = "drop"
  )

# 2. Common theme
fig_theme <- theme_classic(base_size = 14) +
  theme(
    legend.position = "bottom"
  )

# 3. DO plot
p_do <- ggplot(future_annual, aes(x = Year, y = DO, color = ESM)) +
  geom_line(linewidth = 1, alpha = .5) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", linewidth = 0.8) +
  labs(
    title = "",
    x = NULL,
    y = "DO (mg/L)"
  ) +
  scale_color_manual(values = ESM_colors, name = "Model") +
  fig_theme

# 4. pH plot
p_ph <- ggplot(future_annual, aes(x = Year, y = pH, color = ESM)) +
  geom_line(linewidth = 1, alpha = .5) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", linewidth = 0.8) +
  labs(
    title = "",
    x = NULL,
    y = "pH"
  ) +
  scale_color_manual(values = ESM_colors, name = "Model") +
  fig_theme

# 5. Temperature plot
p_temp <- ggplot(future_annual, aes(x = Year, y = Temperature, color = ESM)) +
  geom_line(linewidth = 1, alpha = .5) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", linewidth = 0.8) +
  labs(
    title = "",
    x = "Year",
    y = expression(Temperature~(degree*C))
  ) +
  scale_color_manual(values = ESM_colors, name = "Model") +
  fig_theme

# 5. Upwelling plot
p_upwell <- ggplot(future_annual, aes(x = Year, y = Upwelling, color = ESM)) +
  geom_line(linewidth = 1, alpha = .5) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", linewidth = 0.8) +
  labs(
    title = "",
    x = "Year",
    y = "Upwelling (m2 s-1)") +
  scale_color_manual(values = ESM_colors, name = "Model") +
  fig_theme

# 6. Combine
final_plot <- p_do / p_ph / p_temp / p_upwell +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

final_plot
```

```{r}
# define periods
historical <- future_ocean2 %>% filter(Year <= 2020)
future <- future_ocean2 %>% filter(Year >= 2020)

# summarize
summary_compare <- bind_rows(
  historical %>% mutate(Period = "Historical"),
  future %>% mutate(Period = "Future")
) %>%
  group_by(Period) %>%
  summarize(
    DO = mean(DO_avg, na.rm = TRUE),
    pH = mean(pH_avg, na.rm = TRUE),
    Temp = mean(Temperature_avg, na.rm = TRUE),
    Upwelling = mean(CUTI, na.rm = TRUE)
  )
```

#### Figure 8
```{r}
# 1. Define periods
future_ocean3 <- future_ocean2 %>%
  mutate(
    Period = case_when(
      Year <= 2020 ~ "2000-2020",
      Year >= 2020 ~ "2020-2100",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(Period)) %>% 
  filter(Month %in% c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug"))

# 2. Long format
future_long <- future_ocean3 %>%
  pivot_longer(
    cols = c(DO_avg, pH_avg, Temperature_avg, CUTI),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  mutate(
    Variable = dplyr::recode(Variable,
      DO_avg = "DO",
      pH_avg = "pH",
      Temperature_avg = "Temperature"
    )
  )

# make sure order is correct
future_long$Period <- factor(future_long$Period, levels = c("2000-2020", "2020-2100"))
future_long$Variable <- factor(future_long$Variable, levels = c("DO", "pH", "Temperature", "CUTI"))

fig_theme <- theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

make_boxplot <- function(var_name, y_lab) {
  ggplot(
    dplyr::filter(future_long, Variable == var_name),
    aes(x = Period, y = Value, fill = ESM)
  ) +
    geom_boxplot(
      position = position_dodge(width = 0.75),
      outlier.shape = NA,
      alpha = 0.7
    ) +
    geom_point(
      aes(color = ESM),
      position = position_jitterdodge(
        jitter.width = 0.15,
        dodge.width = 0.75
      ),
      alpha = 0.2,
      size = 0.8
    ) +
    scale_color_manual(values = ESM_colors, name = "Model") +
    scale_fill_manual(values = ESM_colors, name = "Model") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
    labs(
      title = "",
      x = NULL,
      y = y_lab
    ) +
    fig_theme
}

p_do_box <- make_boxplot("DO", "DO (mg/L)")
p_ph_box <- make_boxplot("pH", "pH")
p_temp_box <- make_boxplot("Temperature", expression(Temperature~(degree*C)))
p_upwell_box <- make_boxplot("CUTI", "CUTI (m2/s-1)")

final_boxplot <- (p_do_box | p_ph_box | p_temp_box | p_upwell_box) +
  plot_layout(guides = "collect") &
  theme(legend.position = "")

final_boxplot

ggsave('Figures/fig8.png', width = 8, height = 4, dpi = 300)
```

# Future prediction of recruitment 
#### Prep data 
```{r}
df_future_long <- df_joined %>%
  select(Year, Month, DO_avg, pH_avg, Temperature_avg, CUTI, ESM) %>%
  rename(CUTI_avg = CUTI) %>%
  pivot_longer(
    cols = c(DO_avg, pH_avg, Temperature_avg, CUTI_avg),
    names_to = "driver",
    values_to = "value"
  ) %>%
  mutate(
    Month = factor(Month, levels = month_levels, ordered = TRUE),
    Month_num = match(as.character(Month), month_levels)
  ) %>%
  arrange(ESM, driver, Year, Month_num)
```

#### create rolling windows
```{r}
df_future_windows <- df_future_long %>%
  group_by(ESM, driver, Year) %>%
  arrange(Month_num, .by_group = TRUE) %>%
  mutate(
    value_1mo = value,
    value_3mo = slide_dbl(value, mean, .before = 2, .complete = TRUE),
    value_6mo = slide_dbl(value, mean, .before = 6, .complete = TRUE)
  ) %>%
  ungroup()
```

#### convert to candiate predictors
```{r}
df_future_candidates <- df_future_windows %>%
  pivot_longer(
    cols = starts_with("value_"),
    names_to = "window",
    values_to = "predictor"
  ) %>%
  mutate(
    window = dplyr::recode(
      window,
      value_1mo = "single month",
      value_3mo = "3-month rolling avg",
      value_6mo = "6-month rolling avg"
    )
  ) %>%
  filter(!is.na(predictor))
```

#### join and predict
```{r}
df_future_matched <- df_future_candidates %>%
  inner_join(
    best_models %>% select(driver, Month, window, model),
    by = c("driver", "Month", "window")
  )

future_predictions <- df_future_matched %>%
  group_by(ESM, driver, Month, window) %>%
  mutate(
    pred_obj = map2(model, predictor, ~ {
      as_tibble(predict(.x, newdata = data.frame(predictor = .y), interval = "confidence"))
    })
  ) %>%
  unnest(pred_obj) %>%
  filter(Year >= 2020)
```

#### Figure 7 - timeseries
```{r}

future_predictions_plot <- future_predictions %>%
  mutate(
    date = as.Date(paste(Year, Month, "01", sep = "-"), format = "%Y-%b-%d"),
    driver = dplyr::recode(
      driver,
      DO_avg = "Dissolved Oxygen",
      pH_avg = "pH",
      Temperature_avg = "Temperature",
      CUTI_avg = "CUTI"
    )
  )

fig7 <- ggplot(future_predictions_plot, aes(x = date, y = fit, color = ESM, fill = ESM, group = ESM)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.8, alpha = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.6) +
  facet_wrap(~ driver, scales = "free_y", ncol = 1) +
  scale_color_manual(values = ESM_colors, name = "Model") +
  scale_fill_manual(values = ESM_colors, name = "Model") +
  labs(
    x = "Year",
    y = "Predicted recruitment deviation"
  ) + scale_x_date(
    breaks = seq(as.Date("2020-01-01"), as.Date("2100-01-01"), by = "10 years"),
    date_labels = "%Y"
  ) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom"
  ) +
  theme(legend.position = "")


ggsave("Figures/fig7.png", fig7, width = 8, height = 10, dpi = 300)
```

```{r}
baseline <- df_env_joined %>%
  filter(Year >= 2000, Year <= 2020) %>%
  summarise(mean_dev = mean(dev, na.rm = TRUE)) %>%
  pull(mean_dev)

future_predictions_plot <- future_predictions_plot %>%
  mutate(change = fit - baseline)

ggplot(future_predictions_plot, aes(x = date, y = change, color = ESM)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ driver, scales = "free_y") +
  scale_color_manual(values = ESM_colors) +
  labs(y = "Change in recruitment (relative to 2000–2020)", x = "Year") +
  theme_bw()

ensemble <- future_predictions_plot %>%
  group_by(date, driver) %>%
  summarise(
    mean = mean(fit, na.rm = TRUE),
    sd = sd(fit, na.rm = TRUE),
    .groups = "drop"
  )

ggplot() +
  geom_ribbon(data = ensemble,
              aes(x = date, ymin = mean - sd, ymax = mean + sd),
              fill = "gray70", alpha = 0.5) +

  geom_line(data = ensemble,
            aes(x = date, y = mean),
            color = "black", linewidth = 1.2) +

  geom_line(data = future_predictions_plot,
            aes(x = date, y = fit, color = ESM),
            alpha = 0.5) +

  facet_wrap(~ driver, scales = "free_y") +
  scale_color_manual(values = ESM_colors) +
  theme_bw()
```
```{r}
hist_df <- df_env_joined %>%
  filter(Year >= 2000, Year <= 2020) %>%
  group_by(Year) %>%
  summarise(value = mean(dev, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    Month = NA,
    period = "2000–2020",
    ESM = "Observed"
  )

hist_df <- tidyr::expand_grid(
  hist_df,
  driver = c("DO_avg", "pH_avg", "Temperature_avg", "CUTI_avg")
)

# Future
future_df <- future_predictions %>%
  filter(Year >= 2020) %>%
  mutate(
    period = "2020–2100",
    value = fit
  ) %>%
  select(Year, Month, driver, value, period, ESM)

combined_df <- bind_rows(hist_df, future_df)

combined_df <- bind_rows(hist_df, future_df) %>%
  mutate(
    driver = dplyr::recode(
      driver,
      DO_avg = "Dissolved Oxygen",
      pH_avg = "pH",
      Temperature_avg = "Temperature",
      CUTI_avg = "CUTI"
    )
  )


ggplot(combined_df, aes(x = period, y = value, color = ESM)) +
  
  # boxplots
  geom_boxplot(
    aes(fill = ESM),
    alpha = 0.2,
    outlier.shape = NA,
    position = position_dodge(width = 0.8)
  ) +
  
  geom_jitter(
  alpha = 0.4,
  size = 1,
  position = position_jitterdodge(
    jitter.width = 0.15,
    dodge.width = 0.8
  )
) +
  
  geom_hline(yintercept = 0, linetype = "dashed") +
  
  facet_wrap(~ driver, scales = "fixed") +
  
  scale_color_manual(values = ESM_colors, name = "Model") +
  scale_fill_manual(values = ESM_colors, name = "Model") +
  
  labs(
    x = "",
    y = "Recruitment deviation"
  ) +
  
  theme_bw()
```

```{r}

# Historical annual recruitment
hist_plot_df <- df_env_joined %>%
  filter(Year >= 2000, Year <= 2020) %>%
  group_by(Year) %>%
  summarise(value = mean(dev, na.rm = TRUE), .groups = "drop") %>%
  mutate(group = "Historical")

# Future annual predicted recruitment
future_plot_df <- future_predictions %>%
  filter(Year >= 2020) %>%
  group_by(Year, ESM, driver) %>%
  summarise(value = mean(fit, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    driver = dplyr::recode(
      driver,
      DO_avg = "Dissolved Oxygen",
      pH_avg = "pH",
      Temperature_avg = "Temperature",
      CUTI_avg = "CUTI"
    )
  )

# Shared y-axis limits
y_lims <- range(c(hist_plot_df$value, future_plot_df$value), na.rm = TRUE)

# Historical plot
p_hist <- ggplot(hist_plot_df, aes(x = group, y = value)) +
  geom_boxplot(fill = "gray80", color = "gray30", alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.12, alpha = 0.35, size = 1.2) +  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_cartesian(ylim = y_lims) +
  labs(title = "Historical", x = "", y = "Recruitment deviation") +
  theme_bw()

# Function for future driver plots
make_future_plot <- function(driver_name) {
  ggplot(filter(future_plot_df, driver == driver_name),
         aes(x = ESM, y = value, color = ESM, fill = ESM)) +
    geom_boxplot(alpha = 0.2, outlier.shape = NA) +
    geom_jitter(width = 0.12, alpha = 0.35, size = 1.2) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    scale_color_manual(values = ESM_colors, name = "Model") +
    scale_fill_manual(values = ESM_colors, name = "Model") +
    coord_cartesian(ylim = y_lims) +
    labs(title = driver_name, x = "", y = "") +
    theme_bw() +
    theme(legend.position = "none")
}

p_cuti <- make_future_plot("CUTI")
p_do   <- make_future_plot("Dissolved Oxygen")
p_ph   <- make_future_plot("pH")
p_temp <- make_future_plot("Temperature")


p_hist <- p_hist + theme_bw(base_size = 16) + theme(legend.position = "none")
p_cuti <- p_cuti + theme_bw(base_size = 16) + theme(legend.position = "none")
p_do   <- p_do   + theme_bw(base_size = 16) + theme(legend.position = "none")
p_ph   <- p_ph   + theme_bw(base_size = 16) + theme(legend.position = "none")
p_temp <- p_temp + theme_bw(base_size = 16) + theme(legend.position = "none")


ggsave("hist.png", p_hist, width = 4, height = 4, dpi = 300)
ggsave("cuti.png", p_cuti, width = 4, height = 4, dpi = 300)
ggsave("do.png",   p_do,   width = 4, height = 4, dpi = 300)
ggsave("ph.png",   p_ph,   width = 4, height = 4, dpi = 300)
ggsave("temp.png", p_temp, width = 4, height = 4, dpi = 300)
```

# Correlation between PDO and temperature, pH, and DO (supplemental figure 1)
```{r}
df_all_parameters_march <- df_all_parameters %>%  
  filter(Month == "Mar")

# Pearson correlations
get_cor_info <- function(data, x, y = "mean_pdo") {
  test <- cor.test(data[[x]], data[[y]], method = "pearson", use = "complete.obs")
  
  r <- round(test$estimate, 2)
  pval <- test$p.value
  
  p_label <- ifelse(
    pval < 0.001,
    "p < 0.001",
    paste0("p = ", signif(pval, 2))
  )
  
  list(r = r, pval = pval, p_label = p_label)
}

info_ph <- get_cor_info(df_all_parameters_march, "pH_avg")
info_do <- get_cor_info(df_all_parameters_march, "DO_avg")
info_temp <- get_cor_info(df_all_parameters_march, "Temperature_avg")

# Define base theme
base_theme <- theme_classic(base_size = 14) +
  theme(
    axis.title = element_text(),
    axis.text = element_text(color = "black"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    plot.title = element_text(hjust = 0.5)
  )

# Plot 1: pH vs. PDO
p1 <- ggplot(df_all_parameters_march, aes(x = pH_avg, y = mean_pdo)) +
  geom_point(size = 3, shape = 21, fill = "grey30") +
  geom_smooth(method = "lm", se = FALSE, size = 1.2, color = "black") +
  annotate(
    "text", x = Inf, y = Inf,
    label = paste0("r = ", info_ph$r, "\n", info_ph$p_label),
    hjust = 1.1, vjust = 1.5, size = 5, fontface = "italic", color = "red"
  ) +
  labs(x = "pH", y = "PDO", title = "") +
  base_theme

# Plot 2: DO vs. PDO
p2 <- ggplot(df_all_parameters_march, aes(x = DO_avg, y = mean_pdo)) +
  geom_point(size = 3, shape = 21, fill = "grey30") +
  geom_smooth(method = "lm", se = FALSE, size = 1.2, color = "black") +
  annotate(
    "text", x = Inf, y = Inf,
    label = paste0("r = ", info_do$r, "\n", info_do$p_label),
    hjust = 1.1, vjust = 1.5, size = 5, fontface = "italic", color = "red"
  ) +
  labs(x = "Dissolved Oxygen (mg/L)", y = NULL, title = "") +
  base_theme +
  theme(axis.title.y = element_blank())

# Plot 3: Temperature vs. PDO
p3 <- ggplot(df_all_parameters_march, aes(x = Temperature_avg, y = mean_pdo)) +
  geom_point(size = 3, shape = 21, fill = "grey30") +
  geom_smooth(method = "lm", se = FALSE, size = 1.2, color = "black") +
  annotate(
    "text", x = Inf, y = Inf,
    label = paste0("r = ", info_temp$r, "\n", info_temp$p_label),
    hjust = 1.1, vjust = 1.5, size = 5, fontface = "italic", color = "red"
  ) +
  labs(x = "Temperature (°C)", y = NULL, title = "") +
  base_theme +
  theme(axis.title.y = element_blank())

# Combine with patchwork
(p1 | p2 | p3) +
  plot_annotation(
    theme = theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    )
  )
ggsave("PDO_ROMS_correlation.png", width = 8, height = 4, dpi = 300, bg = "white")

```

